The Weakly Nonlinear Magnetorotational Instability in a thin-gap Taylor Couette flow

### 2016-10-05 paper/thingap_mri_paper.tex
tidied up thingap abstract

\begin{abstract}
The magnetorotational instability (MRI) is a fundamental process of accretion disk physics, but its saturation mechanism remains poorly understood despite considerable theoretical and computational effort. We present a multiple-scales analysis of the non-ideal MRI in the weakly nonlinear regime -- that is, when the the most unstable MRI mode has a growth rate asymptotically approaching zero from above. Here, we develop our theory in a thin-gap, Cartesian channel. Our results confirm Umurhan et al. 2007's findings that the saturation amplitude follows a Ginzburg-Landau equation. We extend these results by performing a detailed force balance for the saturated $u_y$ and $B_z$ fields, demonstrating that even when diffusive effects are important, the bulk flow saturates by a combination of reducing the background shear and rearranging and strengthening the background $B_z$ field. 
\end{abstract}

@@

a real Ginzburg-Landau equation. The saturated solution to Equation \ref{eq:gle} is evidently $\alpha_{saturation} = \pm \sqrt{b/c}$. We plot the first order, second order, and total perturbation structure of the fluid variables in Figures \ref{fig:allorders_velocity} and \ref{fig:allorders_Bfield} with a constant $\alpha_{saturation}$. This is the Ginzburg-Landau equation that was previously found by URM07. Those authors investigated the behavior of this MRI system as a function of $\Pm$. We note that we reproduce the URM07 result that the analytic saturation amplitude scales as $\alpha^2 \propto \Pm^{4/3}$.

@@

The fact that we find a Ginzburg-Landau equation for the amplitude is unsurprising: such an amplitude equation can be found in any system with Euclidean symmetry and a quadratic maximum in growth rate with respect to $k$ \citep{Hoyle:2006}. In this case, the Euclidean symmetry comes from axisymmetry in the $x$-$z$ plane; the quadratic maximum is a consequence of the linear dispersion relation given in Appendix~\ref{app:matrices}.  In Paper II, we show that the same symmetry occurs in the axisymmetric global geometry as well. Figure~\ref{fig:sat_comp_vel} shows the saturated mean radial profiles of $u_0 - u_{y} = -q \Omega_0 x - u_{y}$ and each term in the steady state force balance (i.e. equation~\ref{eqset2} with $\partial_t u_{y} = 0$). In the bulk of the fluid away from the boundary layers, the saturated state shows reduced shear, with little diffusive contribution. This demonstrates that even in a case where diffusive effects are important, the bulk of the fluid saturates by balancing shear and magnetic tension. As discussed at length in \citet{Vasil:2015}, when diffusive effects are not important, it is impossible to rearrange momentum without also rearranging magnetic fields. His model demonstrates saturation without diffusive effects; our results show that outside of the boundary layers, a simultaneous rearrangement of momentum and field occurs. In the boundary layers, the non-linear advection balances viscous dissipation.

>>>>

\begin{abstract}
The magnetorotational instability (MRI) is a fundamental process of accretion disk physics, but its saturation mechanism remains poorly understood despite considerable theoretical and computational effort. We present a multiple-scales analysis of the non-ideal MRI in the weakly nonlinear regime -- that is, when the the most unstable MRI mode has a growth rate asymptotically approaching zero from above. Here, we develop our theory in a thin-gap, Cartesian channel. Our results confirm the finding by Umurhan et al. (2007) that the perturbation amplitude follows a Ginzburg-Landau equation. We extend these results by performing a detailed force balance for the saturated azimuthal velocity and vertical magnetic field, demonstrating that even when diffusive effects are important, the bulk flow saturates via the combined processes of reducing the background shear and rearranging and strengthening the background vertical magnetic field. We directly simulate the Ginzburg-Landau amplitude evolution for our system and demonstrate the pattern formation our model predicts on long length and time scales. 
\end{abstract}

@@

a real Ginzburg-Landau equation. The saturated solution to Equation \ref{eq:gle} is evidently $\alpha_{saturation} = \pm \sqrt{b/c}$. We plot the first order, second order, and total perturbation structure of the fluid variables in Figures \ref{fig:allorders_velocity} and \ref{fig:allorders_Bfield} with a constant $\alpha_{saturation}$. This is the Ginzburg-Landau equation that was previously found by URM07. Those authors investigated the behavior of this MRI system as a function of $\Pm$. We note that we reproduce the URM07 result that the analytic saturation amplitude scales as $\alpha^2 \propto \Pm^{4/3}$ in a thin-gap geometry.

@@

The fact that we find a Ginzburg-Landau equation for the amplitude is unsurprising: such an amplitude equation can be found in any system with Euclidean symmetry and a quadratic maximum in growth rate with respect to the wavenumber \citep{Hoyle:2006}. In this case, the Euclidean symmetry comes from axisymmetry in the $x$-$z$ plane, and the quadratic maximum is a consequence of the linear dispersion relation given in Appendix~\ref{app:dispersion}.  In Paper II, we show that the same symmetry occurs in the axisymmetric global geometry as well.

Here our focus is on a physical description of the saturation mechanism. Figure~\ref{fig:sat_comp_vel} shows the saturated mean radial profiles of $u_0 - u_{y} = -q \Omega_0 x - u_{y}$ and each term in the steady state force balance (i.e. equation~\ref{eqset2} with $\partial_t u_{y} = 0$). In the bulk of the fluid away from the boundary layers, the saturated state shows reduced shear, with little diffusive contribution. This demonstrates that even in a case where diffusive effects are important, the bulk of the fluid saturates by balancing shear and magnetic tension. As discussed at length in \citet{Vasil:2015}, when diffusive effects are not important, it is impossible to rearrange momentum without also rearranging the magnetic field. His model demonstrates saturation without diffusive effects; our results show that outside of the boundary layers, a simultaneous rearrangement of momentum and field occurs. In the boundary layers, the non-linear advection balances viscous dissipation.
